---
title: "Hands-on Exercise 2"
author: "FU YILIN"
---

# 1. Getting Started

In this exercise, we will explore how to enhance \`ggplot2\` visualizations using extensions such as \`ggrepel\`, \`patchwork\`, \`ggthemes\`, and \`hrbrthemes\`.

## **1.1 Installing and loading the required libraries**

We begin by loading the necessary libraries using the \`pacman\` package, which ensures packages are installed before loading them.

```{r}

pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

## **1.2 Importing data**

The dataset Exam_data.csv contains scores for primary school students. We import it using read_csv.

```{r}
exam_data <- read_csv("Data/Exam_data.csv")
```

# **2. Enhancing ggplot2 Annotations with ggrepel**

## 2.1 Base geom_label Annotation 

First, let’s plot English vs. Maths scores using base ggplot2 annotation with geom_label.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## 2.2 Improved Annotation with geom_label_repel

This basic plot uses geom_label, but the labels can overlap. Let’s improve that with geom_label_repel from the ggrepel package.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

# **3. Exploring ggplot2 Themes**

## 3.1 Default Theme

Let’s start with a simple histogram of Maths scores using the default theme_gray.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

## 3.2 Applying theme_economist from ggthemes

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

## 3.3 Applying theme_ipsum from hrbrthemes

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

## 3.4 Customizing theme_ipsum

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

# **4. Beyond Single Graph: Composite Graphics**

We’ll now create multiple plots and combine them using the patchwork package.

## 4.1 Creating the Plots

### 4.1.1 Plot 1: Maths Score Histogram

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```

### 4.1.2 Plot 2: English Score Histogram

```{r}
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```

### 4.1.3 Plot 3: Scatterplot of English vs Maths

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## 4.2 Combining Two Plots Side-by-Side

### 4.2.1 Two Plots Side-by-Side

```{r}
p1 + p2
```

### 4.2.2 Combining Three Plots

```{r}
(p1 / p2) | p3
```

### 4.2.3 Adding Tags to a Composite Figure

```{r}
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

### 4.2.4 Adding an Inset Plot

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### 4.2.5 Composite Plot with Theme

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

This concludes Hands-on Exercise 2. We’ve explored how to annotate, theme, and combine ggplot2 graphics using several powerful R packages.
