---
title: "Take-home Exercise 1 - Understanding Singapore's Demographic Landscape: Population Age Structure and Spatial Patterns in 2024"
author: "FU YILIN"
---

# Part 1:

# 1 Overview

## 1.1 Background

Singapore’s demographic profile provides critical insights for policy planning, urban development, and resource allocation. The dataset `respopagesex2024.csv` contains population counts by age, sex, planning area, and subzone, based on Singapore's official population statistics for 2024. This research aims to uncover key trends in aging and spatial demographic variations using visual analytics.

## 1.2 The Data

The dataset used is `respopagesex2024.csv` obtained from the Department of Statistics Singapore. It includes the following fields:

-   PA: Planning Area

-   SZ: Subzone

-   Age: Age in years (0 to 90+)

-   Sex: Male or Female

-   Pop: Resident count

-   Time: Time / Period

::: {.callout-note title="Note" icon="false"}
```         
1) For June 2024, Planning Areas refer to areas demarcated in the Urban Redevelopment Authority's Master Plan 2019. 
2) Data from 2003 onwards exclude residents who have been away from Singapore for a continuous period of 12 months or longer as at the reference period. 
3) The figures have been rounded to the nearest 10. 
4) The data may not add up due to rounding.
```
:::

## 1.3 Research Objectives

Using the `Singapore Residents by Planning Area / Subzone, Single Year of Age and Sex, June 2024` dataset, this article aims to use appropriate Exploratory Data Analysis (EDA) methods and `ggplot2` function to explore:

-   Areas with pronounced aging patterns and low birth rates

-   Spatial variation of age distribution across planning areas and regions

# 2 Loading Packages

The following packages are used for data wrangling and visualization:

-   `tidyverse`: A collection of R packages for data science, including `dplyr`, `ggplot2`, and others for data manipulation and visualization.

-   `ggplot2`: A grammar of graphics for creating static, animated, and interactive data visualizations.

-   `dplyr`: Provides a set of functions for efficiently manipulating datasets (e.g., filtering, selecting, and summarising).

-   `forcats`: Tools for working with categorical variables (factors), especially useful for reordering and relabeling.

-   `ggthemes`: Offers additional themes and scales to improve the aesthetics of `ggplot2` charts.

-   `patchwork`: Enables combining multiple `ggplot2` plots into a single layout.

-   `knitr`: Facilitates dynamic report generation by weaving R code into documents.

-   `gt`: Used to create elegant tables for display in HTML or markdown reports.

The following code chunk uses `p_load ()` of `pacman` packages to check if `tidyverse` packages are installed and can be called into R.

```{r}
pacman::p_load(tidyverse, ggplot2, dplyr, forcats, ggthemes, patchwork, knitr, gt)
```

# 3 Data Preparation

In this section, we will wrangle the data to prepare for analysis and visualization.

## 3.1 Dataset Overview

::: {.callout-note title="Note"}
This dataset stores "90_and_Over" as a character value in the Age column, we need to clean and convert it to a numeric value to perform age-based analysis. We use 95 as a conservative numeric replacement.
:::

```{r}
data <- read_csv("Data/respopagesex2024.csv") %>%
  mutate(Age = ifelse(Age == "90_and_Over", "95", Age),
         Age = as.numeric(Age)) %>%
  drop_na(Age)
```

The dataset consists of 60,424 rows and 6 columns, representing Singapore’s resident population as of 2024.

A warning was raised when converting the Age column from character to numeric:

::: {.callout-warning title="Warning"}
NAs introduced by coercion
:::

This indicates that at least one value in the Age column was non-numeric (e.g., "Total" or "Unknown"), which cannot be coerced to a number and was converted to NA. To ensure accurate analysis, those invalid rows were excluded using `drop_na(Age)`. This step guarantees that only valid age-based population entries are used in further calculations.

## 3.2 Duplicate Check

```{r}
sum(duplicated(data))
data[duplicated(data), ]
```

Result suggests that there are no fully duplicate rows in this dataset.

After cleaning, 60,424 valid records remain for analysis. No duplicate entries were detected. Non-numeric age entries, likely representing summary rows such as ‘Total’, were excluded.

## 3.3 Subzone Summary

The code chunk below creates a summary table that shows all subzones belonging to each planning area, displayed cleanly using the `gt()` package for better readability. It’s useful for understanding the structure and granularity of the geographic data.

```{r}
grouped <- data %>%
  select(PA, SZ) %>%
  distinct() %>%
  arrange(PA, SZ) %>%
  group_by(PA) %>%
  summarise(Subzones = paste(SZ, collapse = ", "))

grouped %>%
  gt() %>%
  tab_header(
    title = "Planning Areas and Their Subzones"
  )
```

# 4 Exploratory Data Analysis

## 4.1 Age-Sex Pyramid

This population pyramid visualizes the age and gender distribution. Males are plotted on the left and females on the right, highlighting the dependency load and gender imbalance at older ages.

```{r}
pyramid_data <- data %>%
  group_by(Age, Sex) %>%
  summarise(Pop = sum(Pop), .groups = "drop") %>%
  mutate(Pop = if_else(Sex == "Males", -Pop, Pop))

pyramid_plot <- ggplot(pyramid_data, aes(x = Age, y = Pop, fill = Sex)) +
  geom_bar(stat = "identity", width = 1) +
  coord_flip() +
  scale_y_continuous(labels = abs) +
  labs(title = "Singapore Resident Population Pyramid (2024)",
       x = "Age", y = "Population") +
  theme_minimal()

pyramid_plot
```

**Observations**:

-   A bulge around ages 30–50 suggests a dominant working-age population.

-   Females dominate in ages 80+, reflecting higher life expectancy.

-   Lower base suggests declining birth rates.

## 4.2 Aging Structure Across Planning Areas

To better understand the spatial dynamics of population aging, we compared the total population of the **young group (aged 0–14)** and the **aging group (aged 65 and above)** across Singapore’s Planning Areas (PAs). This analysis highlights demographic imbalances by region, identifying areas where the elderly population significantly exceeds the youth population, and vice versa.

```{r,fig.width=12, fig.height=8}
age_groups <- data %>%
  filter(Age <= 14 | Age >= 65) %>%
  mutate(group = case_when(
    Age <= 14 ~ "Young",
    Age >= 65 ~ "Aging"
  )) %>%
  group_by(PA, group) %>%
  summarise(total_pop = sum(Pop), .groups = "drop")

ggplot(age_groups, aes(x = reorder(PA, -total_pop), y = total_pop, fill = group)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Young vs Aging Population by Planning Area",
       x = "Planning Area", y = "Population", fill = "Age Group") +
  theme_minimal()
```

We refine the visualization by focusing on the top 20 Planning Areas with the highest combined population of young (aged 0–14) and elderly (aged 65 and above) residents. The `top_pa` object identifies these most populous areas based on total demographic burden, while `top_age_groups` filters the dataset accordingly to ensure that only these selected Planning Areas are included in the plot for clearer comparison.

```{r,fig.width=12, fig.height=8}
top_pa <- age_groups %>%
  group_by(PA) %>%
  summarise(total = sum(total_pop)) %>%
  arrange(desc(total)) %>%
  slice_head(n = 20)  # select top 20

top_age_groups <- age_groups %>%
  filter(PA %in% top_pa$PA)

# Define the order (e.g. by total population descending)
pa_order <- top_age_groups %>%
  group_by(PA) %>%
  summarise(total = sum(total_pop)) %>%
  arrange(desc(total)) %>%
  pull(PA)

# Apply the same factor levels
top_age_groups <- top_age_groups %>%
  mutate(PA = factor(PA, levels = pa_order))

ratio_data <- age_groups %>%
  filter(PA %in% top_pa$PA) %>%
  pivot_wider(names_from = group, values_from = total_pop) %>%
  mutate(ratio = Aging / Young) %>%
  filter(!is.na(PA)) %>%  # Remove NAs
  mutate(PA = factor(PA, levels = pa_order))

# Top bar chart
bar_plot <- ggplot(top_age_groups, aes(x = total_pop, y = fct_rev(PA), fill = group)) +
  geom_col(position = "dodge") +
  labs(x = "Population", y = NULL, fill = "Age Group") +
  theme_minimal()

# Bottom line chart with consistent PA order
line_plot <- ggplot(ratio_data, aes(x = PA, y = ratio)) +
  geom_point(size = 3, color = "black") +
  geom_line(aes(group = 1), color = "black") +
  labs(x = "Planning Area", y = "Aging / Young Ratio") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine them
(bar_plot / line_plot) + plot_layout(heights = c(2, 1)) +
  plot_annotation(title = "Population Distribution and Aging Ratio by Planning Area")

```

::: {.callout-tip title="Tip"}
To enlarge the graph for better readability, we can increase the figure width and height:

-   `fig.width=12` → makes the chart wider
-   `fig.height=8` → gives enough vertical space for stacked or combined plots

Both plots use consistent PA ordering for easier comparison.
:::

**Aging-Heavy Areas**: these planning areas have a high Aging-to-Young ratio, meaning significantly more elderly (65+) than young residents (0–14):

-   Ang Mo Kio has the highest aging ratio (above 2.0) despite moderate total population — indicating a strongly aging population.

-   Bukit Merah and Queenstown also exhibit high aging ratios with noticeably lower young population bars. These are classic mature estates with older residents aging in place.

-   Bedok and Toa Payoh show both large elderly populations and high aging ratios.

**Youth-Dense Areas**:

-   Punggol has the lowest aging ratio (under 0.5) and a visibly dominant young population bar — a clear sign of a family-oriented, younger estate.

-   Sengkang and Tampines also have low aging ratios and high young populations, suggesting they are popular for younger families.

-   Jurong West and Woodlands show balanced profiles but still lean younger compared to the national median.

## 4.3 Age Distribution by Region Type

To examine regional differences in demographic structure, a boxplot was generated to illustrate the distribution of resident ages across Singapore’s various region types, as defined by their respective Planning Areas.

The Urban Authority of Singapore (URA) has categorized Singapore into 5 regions for urban planning purposes, namely Central Region, North Region, North-East Region, East Region and West Region.

By aligning Planning Areas with these official region classifications, the analysis enables a clearer understanding of age-related demographic patterns across the island.

```{r, fig.width=14, fig.height=10}
expanded_data <- data %>%
  filter(!is.na(PA)) %>%
  mutate(Age = ifelse(Age == "90_and_Over", "95", Age),
         Age = as.numeric(Age)) %>%
  filter(!is.na(Age)) %>%
  mutate(PA = str_trim(PA)) %>%
  mutate(region_type = case_when(
    # Core Central Region
    PA %in% c("Downtown Core", "Outram", "Sentosa", "Rochor", "Orchard", "Newton",
              "River Valley", "Bukit Timah", "Holland Road", "Tanglin", "Novena", "Thomson") ~ "Core Central Region",
    
    # Rest of Central Region
    PA %in% c("Bishan", "Bukit Merah", "Geylang", "Kallang", "Marine Parade", "Queenstown", "Southern Islands", "Toa Payoh") ~ "Rest of Central Region",
    
    # North Region
    PA %in% c("Central Water Catchment", "Lim Chu Kang", "Mandai", "Sembawang", "Simpang", "Sungei Kadut", "Woodlands", "Yishun") ~ "North Region",
    
    # North-East Region
    PA %in% c("Ang Mo Kio", "Hougang", "North-Eastern Islands", "Punggol", "Seletar",
              "Sengkang", "Serangoon") ~ "North-East Region",
    
    # East Region
    PA %in% c("Bedok", "Changi", "Changi Bay", "Paya Lebar", "Pasir Ris", "Tampines") ~ "East Region",
    
    # West Region
    PA %in% c("Bukit Batok", "Bukit Panjang", "Boon Lay", "Pioneer", "Choa Chu Kang", "Clementi", "Jurong East", "Jurong West", "Tengah", "Tuas", "Western Islands", "Western Water Catchment") ~ "West Region",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(region_type)) %>%
  uncount(weights = Pop)
  
p1 <- ggplot(expanded_data, aes(x = region_type, y = Age, fill = region_type)) +
  geom_boxplot() +
  labs(title = "Age Distribution by Region Type", x = "Region", y = "Age") +
  theme_minimal() +
  coord_flip()

p2 <- ggplot(expanded_data, aes(x = Age, y = fct_reorder(PA, Age), fill = region_type)) +
  geom_boxplot() +
  facet_wrap(~ region_type, scales = "free_y") +
  labs(
    title = "Age Distribution by Planning Area and Region",
    x = "Age",
    y = "Planning Area"
  ) +
  theme_minimal()

p1 / p2 +
  plot_annotation(title = "Overview: Regional and Planning Area Age Distribution")
```

From **Age Distribution by Region Type** figure, we can observe that overall age distribution is similar. All regions have interquartile ranges (IQRs) roughly between 25 and 65 years. Core Central Region has a slightly wider IQR; median age around mid-40s. East Region has a slightly lower median than others. North Region & North-East Region have similar median and IQR (\~25–65), slightly lower upper quartile, possibly reflects more family-centric, middle-aged population. West Region also has a broad IQR with no extreme values. No region shows a highly youth-dominant or elderly-dominant skew in isolation.

From the **Age Distribution by Planing Area and Region** figure, we can observe notable outlines in the plots.

-   North-East Region – Seletar: Has multiple high outliers (above age 80+), despite a narrow IQR.

-   West Region – Tengah: Displays a string of high outliers, unusually concentrated near 85–90. Tengah is a newly developing area, so this is surprising.

-   Core Central Region – Downtown Core, Outram, Newton: Outliers at both very low and very high ages, which reflects age-diverse.

# 5 Summary

Singapore's 2024 demographic landscape is defined by a mature population with a strong working-age core, declining birth rates, and regional variance in aging patterns. Aging is concentrated in older housing estates, while younger families cluster in newer towns like Punggol and Sengkang.

# 6 References

-   Department of Statistics Singapore. (2024). Resident Population by Age, Sex and Planning Area/Subzone. https://www.singstat.gov.sg/
-   Wikipedia contributors. (2024). Regions of Singapore. [https://en.wikipedia.org/wiki/Regions_of_Singaporehttps://en.wikipedia.org/wiki/Regions_of_Singapore](https://en.wikipedia.org/wiki/Regions_of_Singapore)

# Part 2: Learn from Peers

Reference: [Peer's Submission](https://isss-608-yihao-yuan.netlify.app/take-home_ex/take-home_ex01/take-home_ex01#population-by-age-group-and-planning-area-analyze-the-number-of-the-aging-population-in-each-region.)

## 1 Original Submission

```{r}

res_data_grouped <- data %>%
  mutate(AgeGroup = case_when(
    Age >= 0 & Age <= 20 ~ "Under Age",
    Age >= 21 & Age <= 60 ~ "Adult",
    Age >= 61 ~ "Aging"
  )) %>%
  group_by(PA, AgeGroup) %>%
  summarise(Pop = sum(Pop, na.rm = TRUE), .groups = "drop")

res_data_grouped_filtered <- res_data_grouped %>%
  filter(AgeGroup != "Adult")

age_summary <- res_data_grouped %>%
  filter(AgeGroup %in% c("Under Age", "Aging")) %>%
  group_by(PA, AgeGroup) %>%
  summarise(Total = sum(Pop), .groups = "drop")


age_ratio <- age_summary %>%
  tidyr::pivot_wider(names_from = AgeGroup, values_from = Total) %>%
  mutate(ratio = Aging / `Under Age`)

ggplot(age_ratio, aes(x = ratio, y = reorder(PA, ratio))) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Aging-to-Under-Age Population Ratio by Planning Area",
    x = "Aging / Under Age Ratio",
    y = "Planning Area"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 5),
    plot.title = element_text(face = "bold")
  )
```

## **2 Three Good Design Principles Applied**

-   **Clear Title and Labels**

    -   The title clearly describes the data being presented: “Aging-to-Under-Age Population Ratio by Planning Area.”

    -   Axis labels are descriptive, making the visual easily understandable.

-   **Horizontal Bar Chart for Easy Comparison**

    -   The use of a horizontal bar chart effectively displays the aging-to-under-age ratio by planning area, allowing easy comparison across areas.

-   **Sorted Order for Better Analysis**

    -   Planning areas are sorted by the aging-to-under-age ratio, making it easy to identify areas with the highest and lowest ratios.

## **3 Three Areas for Further Improvement**

-   **Enhanced Visualization**

    A gradient color scheme from light blue (low ratio) to dark red (high ratio) enhances visual clarity, showing which areas have a higher aging population.

-   **Optimized Annotations**

    -   A dashed vertical line at ratio = 1 indicates a balanced population between the aging and under-age groups.

    -   The highest and lowest ratios are directly annotated, making it easy to identify critical areas.

-   **Filter Areas with No Values**

    Areas with zero values (no data) have been removed from the visualization, ensuring only meaningful data is displayed.

## **4 Makeover Version**

```{r,fig.width=12, fig.height=8}
# Grouping Data into Age Categories
data <- data %>%
  mutate(AgeGroup = case_when(
    Age >= 0 & Age <= 20 ~ "Under Age",
    Age >= 21 & Age <= 60 ~ "Adult",
    Age >= 61 ~ "Aging"
  ))

# Calculating Population by Age Group for Each Planning Area
res_data_grouped <- data %>%
  group_by(PA, AgeGroup) %>%
  summarise(Pop = sum(Pop, na.rm = TRUE), .groups = "drop")

# Filtering Only "Under Age" and "Aging"
res_data_filtered <- res_data_grouped %>%
  filter(AgeGroup != "Adult")

# Pivoting Data to Calculate Aging-to-Under-Age Ratio
age_summary <- res_data_filtered %>%
  pivot_wider(names_from = AgeGroup, values_from = Pop, values_fill = 0) %>%
  mutate(ratio = Aging / `Under Age`) %>%
  filter(!is.na(ratio))

# Sorting by Ratio for Better Visualization
age_summary <- age_summary %>% arrange(ratio)

# Calculate Dynamic Plot Height (1.2 times number of planning areas)
plot_height <- nrow(age_summary) * 0.3

# Create the Enhanced Plot
p <- ggplot(age_summary, aes(x = ratio, y = reorder(PA, ratio), fill = ratio)) +
  geom_col() +
  scale_fill_gradient(low = "lightblue", high = "darkred") +
  geom_vline(xintercept = 1, color = "gray", linetype = "dashed") +
  geom_text(aes(label = ifelse(ratio == max(ratio, na.rm = TRUE), 
                               paste("Highest:", round(ratio, 2)), 
                               ifelse(ratio == min(ratio, na.rm = TRUE) & ratio > 0, 
                                      paste("Lowest:", round(ratio, 2)), 
                                      ""))), 
            hjust = -0.1, size = 3.5, color = "black") +
  labs(
    title = "Enhanced Aging-to-Under-Age Population Ratio by Planning Area",
    x = "Aging / Under Age Ratio",
    y = "Planning Area"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.y = element_text(size = 10, hjust = 1, margin = margin(r = 8)),
    axis.text.x = element_text(size = 10),
    plot.margin = margin(1, 2, 1, 6, "cm"),
    legend.position = "none"
  ) +
  coord_cartesian(clip = "off", ylim = c(1, nrow(age_summary)))

# Display the Plot
p

```

::: {.callout-note title="Note"}
**`mutate(ratio = Aging / Under Age)`:** Calculates the ratio of the aging population to the under-age population for each planning area.

**`filter(!is.na(ratio))`:** Removes any rows where the ratio calculation is not possible (e.g., missing values).
:::
